---
title: "Introduction to rminka"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rminka}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # para CRAN pon eval=FALSE en los chunks que llaman a la API
  # de momento lo dejamos TRUE para que veas que funciona
  eval = TRUE
)

library(dplyr)
library(tibble)
library(rminka)
library(sf)
library(leaflet)
library(htmltools)
#library(magick)

```

```{r leaflet-css, echo=FALSE, results='asis'}
# solve CSS errors
cat('
<style>
.leaflet-container img.leaflet-marker-icon,
.leaflet-container img.leaflet-marker-shadow {
  max-width: none !important;
  max-height: none !important;
  background: transparent !important;
  border: none !important;
  box-shadow: none !important;
}
.leaflet-popup-content img {
  max-width: 250px !important;
  height: auto !important;
  display: block;
  margin: 0 auto;
}
</style>
')
```

This document introduces you to rminka basic set of tools, and shows you how to 
apply them to obtain your desired information.
Once you’ve installed, read vignette("rminka") to learn more.


### - Project Queries 

Project-related functions will be illustrated using the project Biomarató Tarragona 2025.

- **<u>mnk_proj_byname</u>**

Initially, only the project name is known, so a search is performed to retrieve the corresponding project ID. This is done with mnk_proj_byname(). Here we use the query "biomarato 2025".

```{r project1, include = TRUE , echo=TRUE}
prj_names <- mnk_proj_byname("biomarato 2025")

prj_names 

# For this example 

prj_names[,c(1:2)]

```
- **<u>mnk_proj_info</u>**
 
Once the project ID is known, detailed information can be retrieved with mnk_proj_info(). For the Biomarató Tarragona 2025, the project ID is 419.

```{r project_info, include = TRUE , echo=TRUE}

prj_info <- mnk_proj_info(419)

prj_info

```
- **<u>mnk_proj_user</u>**

Users explicitly subscribed to a project can be retrieved with mnk_proj_user() using the project ID

```{r project_info_user, include = TRUE , echo=TRUE}

prj_user <- mnk_proj_user(419)

prj_user

```
Note: The simplest way to retrieve all participants in a project — including those who are not subscribed — is to download the project's observations with `mnk_proj_obs()` and then extract the unique values of user_login. Each observation includes the observer's login, so the set of distinct logins corresponds to the project's participants.

If the project lasts only one year, or if you are only interested in the participants from a single month, a single function call is sufficient. For example, project 419 ran in 2025, so all its observations can be obtained with `mnk_proj_obs(419, year = 2025)`:

```{r project_info_user2, include = TRUE , echo=TRUE, eval=TRUE}

library(rminka)
library(dplyr)

# 1. Download all observations for the project in may 2025

obs_2025 <- mnk_proj_obs(419,year=2025, mont=5, quiet = TRUE)

# 2. Extract the unique participants

participants_may_2025 <- obs_2025 %>%
  distinct(user_login) %>%
  arrange(user_login)

participants_may_2025

```


  - **<u>mnk_proj_obs</u>**

Returns all observations for a project within a selected year, and optionally within a specific month. For the Biomarató Tarragona 2025 (project ID 419), observations for the entire year are obtained with `mnk_proj_obs(419, year = 2025)`, while observations for a single month — for example, May — are obtained with `mnk_proj_obs(419, year = 2025, month = 5)`.

```{r project_obs:proj, include = TRUE , echo=TRUE}

prj_obs <- mnk_proj_obs(419, year= 2025, month=5)

prj_obs[2:14]

```


### - User Queries  

User-related functions will be illustrated using the user Xavier Salvador.
  
 - **<u>mnk_user_byname</u>**
 
Initially, only an approximate name is known, so a search is performed to retrieve the corresponding user_login. For this example, we start with the query "xavi", which returns a list of possible logins. The desired user — Xavier Salvador — is then selected from that list.

```{r user_name, include = TRUE , echo=TRUE}

user_name <- mnk_user_byname("xavi")

user_name

```


- **<u>mnk_user_info</u>**
 
Once the user ID is known, detailed information can be retrieved with mnk_user_info(). For Xavier Salvador (login "xasalva"), the user ID is 4
   
```{r user_info, include = TRUE , echo=TRUE}

user_info <- mnk_user_info(4)

user_info

```


 - **<u>mnk_user_proj</u>**
 
 mnk_user_proj() returns the projects to which a user is explicitly subscribed, given the user ID. For Xavier Salvador (user ID 4), the list is obtained with `mnk_user_proj(4)`.

Note: Projects in which a user has contributed observations but is not formally subscribed cannot be retrieved directly in R.
 
```{r user_project, include = TRUE , echo=TRUE}

user_project <- mnk_user_proj(4)

user_project

```


 - **<u> mnk_user_obs</u>**
 
Returns all observations for a user within a selected year, and optionally within a specific month. For Xavier Salvador (user ID 4), observations for the entire year are obtained with `mnk_user_obs(4, year = 2025)`, while observations for a single month, for example May, are obtained with `mnk_user_obs(4, year = 2025, month = 5)`.

```{r user_obs, include = TRUE , echo=TRUE}

user_obs <- mnk_user_obs(user_id= 4, year = 2025, month = 8)

user_obs

```


### - Place Queries

Place-related functions will be illustrated using the place  Piscines del Fòrum
 
 - **<u>mnk_place_byname</u>**
 
Initially, only the place name ( but not exactly) is known, so a search will be performed to find the place ID

```{r place_name, include = TRUE , echo=TRUE}
library(rminka)

places <- mnk_place_byname("Forum")
places[,1:6]

```

 - **<u>mnk_place_sf</u>**
 

```{r place_sf, include = TRUE , echo=TRUE}

library(sf)
library(leaflet)


# 1. Downloading the geometry 
place <- mnk_place_sf(253)

# 2. Drawing the map
 forum_sf <-leaflet(place) |>
                addProviderTiles("OpenStreetMap", group = "OSM") |>
                addProviderTiles("Esri.WorldImagery", group = "Satélite") |>
                addPolygons(
                  color = "#2c4fb8",
                  weight = 2,
                  opacity = 1,
                  fillOpacity = 0.4,
                  label = ~name, # information added from previous function
                  highlightOptions = highlightOptions(weight = 3, bringToFront = TRUE)
                ) |>
                addLayersControl(baseGroups = c("Satélite", "OSM")) 
 forum_sf

```

 - **<u>mnk_places_obs</u>**
 
```{r places_obs, include = TRUE , echo=TRUE}

obs_place <- mnk_place_obs(place_id = 253, year = 2025, month = 2, quiet = TRUE)

obs_place

#Turning the dataframe into an sf object with the mnk_obs_sf() function

obs_sf <- mnk_obs_sf(obs_place,"id","taxon_name","observed_on", "url_picture", "uri", "user_login")

# Preparing the obtained data for later display in the marker popup

popup_final <-  paste0( "ID: <a href='", obs_sf$uri, "' target='_blank'>",obs_sf$id , "</a><br>",
                        "Specie:", obs_sf$taxon_name, "<br>",
                        "Observer: ", obs_sf$user_login,"<br>",
                        "Date:", obs_sf$observed_on, "<br>",
                        "<a href= '", obs_sf$url_picture, "' target='_blank'><img src='", 
                        obs_sf$url_picture, "' style='margin-top:2px;border-radius:4px;'> </a> ")
#finally plotins

leaflet(obs_sf) %>%
  addTiles() %>%
  addMarkers(lng = ~longitude,
             lat = ~latitude,
             popup = popup_final)


 forum_sf |>
  addMarkers(data = obs_sf, popup = ~popup_final)

```


### - Observation Queries

 - **<u>mnk_obs_id</u>**

```{r obs_id, include = TRUE , echo=TRUE}

obs_id <- mnk_obs_id(id = 553028)

obs_id

```

 - **<u>mnk_obs</u>**
 
```{r obs, include = TRUE , echo=TRUE}

#In this example don´t show messages in console (quiet= TRUE)

obs <- mnk_obs(taxon_name= "Diplodus sargus", year=2025, user_id=6,quiet = TRUE)

obs[,c(1,21,2,10,11,16,27)]

obs


```

 - **<u>mnk_obs_byday</u>**
 
 ### - Auxiliary functions

rminka includes helper functions that support the main queries. These include tools to convert observation tables into sf objects for mapping, to resolve taxonomic information, and to prepare data for visualization. They are designed to connect API results directly with standard R workflows for analysis and reporting.
 

