---
title: "Introduction to rminka"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rminka}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # para CRAN pon eval=FALSE en los chunks que llaman a la API
  # de momento lo dejamos TRUE para que veas que funciona
  eval = TRUE
)

library(dplyr)
library(tibble)
library(rminka)
library(sf)
library(leaflet)
library(htmltools)
#library(magick)

```

```{r leaflet-css, echo=FALSE, results='asis'}
# solve CSS errors
cat('
<style>
.leaflet-container img.leaflet-marker-icon,
.leaflet-container img.leaflet-marker-shadow {
  max-width: none !important;
  max-height: none !important;
  background: transparent !important;
  border: none !important;
  box-shadow: none !important;
}
.leaflet-popup-content img {
  max-width: 250px !important;
  height: auto !important;
  display: block;
  margin: 0 auto;
}
</style>
')
```

This document introduces you to rminka basic set of tools, and shows you how to 
apply them to obtain your desired information.
Once you’ve installed, read vignette("rminka") to learn more.

We will develop an example based on a project: Biomarato Tarragona 2025. 
The example will focus on this project and one of its participants, Xavier Salvador. 
As the example progresses, we will use the different types of functions from the package.


### - Project Queries 

- **<u>mnk_proj_byname</u>**

Initially, only the project name is known, so a search will be performed to find 
the project ID (project_id). The `mnk_proj_byname` function will be used for this purpose. 
A generic search will be conducted using the text "biomarato 2025".



```{r project1, include = TRUE , echo=TRUE}
prj_names <- mnk_proj_byname("biomarato 2025")

prj_names 

# For this example 

prj_names[,c(1:2)]

```
- **<u>mnk_proj_info</u>**
 
   Retrieves detailed project information using its known ID.
```{r project_info, include = TRUE , echo=TRUE}

prj_info <- mnk_proj_info(419)

prj_info

```
- **<u>mnk_proj_user</u>**

It´s posible to known the users explicited subscribed in this projects using 
this function

```{r project_info_user, include = TRUE , echo=TRUE}

prj_user <- mnk_proj_user(419)

prj_user

```

  - **<u>mnk_proj_obs</u>**

Retrieves 

```{r project_obs:proj, include = TRUE , echo=TRUE}

prj_obs <- mnk_proj_obs(419, year= 2025, month=5)

prj_obs[2:14]

```


### - User Queries  
  
 - **<u>mnk_user_byname</u>**

```{r user_name, include = TRUE , echo=TRUE}

user_name <- mnk_user_byname("xavi")

user_name

```


- **<u>mnk_proj_info</u>**
 
   Retrieves detailed project information using its known ID.
   
```{r user_info, include = TRUE , echo=TRUE}

user_info <- mnk_user_info(4)

user_info

```


 - **<u>mnk_user_proj</u>**
 
```{r user_project, include = TRUE , echo=TRUE}

user_project <- mnk_user_proj(4)

user_project

```


 - **<u> mnk_user_obs</u>**

```{r user_obs, include = TRUE , echo=TRUE}

user_obs <- mnk_user_obs(user_id= 4, year = 2025, month = 8)

user_obs

```


### - Place Queries

 - **<u>mnk_place_byname</u>**

```{r place_name, include = TRUE , echo=TRUE}
library(rminka)

places <- mnk_place_byname("Forum")
places[,1:6]

```

 - **<u>mnk_place_sf</u>**
 

```{r place_sf, include = TRUE , echo=TRUE}

library(sf)
library(leaflet)


# 1. Downloading the geometry 
place <- mnk_place_sf(253)

# 2. Drawing the map
 forum_sf <-leaflet(place) |>
                addProviderTiles("OpenStreetMap", group = "OSM") |>
                addProviderTiles("Esri.WorldImagery", group = "Satélite") |>
                addPolygons(
                  color = "#2c4fb8",
                  weight = 2,
                  opacity = 1,
                  fillOpacity = 0.4,
                  label = ~name, # information added from previous function
                  highlightOptions = highlightOptions(weight = 3, bringToFront = TRUE)
                ) |>
                addLayersControl(baseGroups = c("Satélite", "OSM")) 
 forum_sf

```

 - **<u>mnk_places_obs</u>**
 
```{r places_obs, include = TRUE , echo=TRUE}

obs_place <- mnk_place_obs(place_id = 253, year = 2025, month = 2)

obs_place

#Turning the dataframe into an sf object with the mnk_obs_sf() function

obs_sf <- mnk_obs_sf(obs_place,"id","taxon_name","observed_on", "url_picture", "uri", "user_login")

# Preparing the obtained data for later display in the marker popup
