---
title: "Introduction to rminka"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rminka}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # para CRAN pon eval=FALSE en los chunks que llaman a la API
  # de momento lo dejamos TRUE para que veas que funciona
  eval = TRUE,
  fig.align = "center",   # centered
  out.width = "600px"   # fixed size
  )

library(dplyr)
library(tibble)
library(rminka)
library(sf)
library(leaflet)
library(htmltools)
library(magick)

```

```{r leaflet-css, echo=FALSE, results='asis'} 

cat('  
<style>
/*  solve css error marker */
.leaflet-container img.leaflet-marker-icon,
.leaflet-container img.leaflet-marker-shadow {
  max-width: none !important;
  max-height: none !important;
  background: transparent !important;
  border: none !important;
  box-shadow: none !important;
}
.leaflet-popup-content img {
  max-width: 100px !important;
  height: auto !important;
  display: block;
  margin: 0 auto;
}

/* leaflet center and size  */
.leaflet, .html-widget {
  width:100% !important;
  height: 60vh !important;   /* 60% de la altura de la ventana */
  max-height: 500px;    
  margin-left: auto !important;
  margin-right: auto !important;
  display: block !important;
}

/* image center and size  */
img {
  display: block;
  margin-left: auto !important;
  margin-right: auto !important;
  max-width: 900px !important;
  height: auto !important;
}
</style>
')

```
This document introduces you to rminka basic set of tools, and shows you how to 
apply them to obtain your desired information.
Once you’ve installed, read vignette("rminka") to learn more.


### <u>- Project Queries </u>

Project-related functions will be illustrated using the project Biomarató Tarragona 2025.

### ● `mnk_proj_byname()`

Initially, only the project name is known, so a search is performed to retrieve the corresponding project ID. This is done with `mnk_proj_byname()`. Here we use the query "biomarato 2025".

```{r project1, include = TRUE , echo=TRUE}
prj_names <- mnk_proj_byname("biomarato 2025")

prj_names 

# For this example 

prj_names[,c(1:2)]

```
### ● `mnk_proj_info()`
 
Once the project ID is known, detailed information can be retrieved with `mnk_proj_info()`. For the Biomarató Tarragona 2025, the project ID is 419.

```{r project_info, include = TRUE , echo=TRUE}
prj_info <- mnk_proj_info(419)

prj_info

```
### ● `mnk_proj_user()`

Users explicitly subscribed to a project can be retrieved with `mnk_proj_user()` using the project ID

```{r project_info_user, include = TRUE , echo=TRUE}
prj_user <- mnk_proj_user(419)

prj_user

```
Note: The simplest way to retrieve all participants in a project — including those who are not subscribed — is to download the project's observations with `mnk_proj_obs()` and then extract the unique values of user_login. Each observation includes the observer's login, so the set of distinct logins corresponds to the project's participants.

If the project lasts only one year, or if you are only interested in the participants from a single month, a single function call is sufficient. For example, project 419 ran in 2025, so all its observations can be obtained with `mnk_proj_obs(419, year = 2025)`:

```{r project_info_user2, include = TRUE , echo=TRUE, eval=TRUE}
# 1. Download all observations for the project in may 2025

obs_2025 <- mnk_proj_obs(419,year=2025, mont=5, quiet = TRUE)

# 2. Extract the unique participants

participants_may_2025 <- obs_2025 %>%
  distinct(user_login) %>%
  arrange(user_login)

participants_may_2025

```


### ● `mnk_proj_obs()`

Returns all observations for a project within a selected year, and optionally within a specific month. For the Biomarató Tarragona 2025 (project ID 419), observations for the entire year are obtained with `mnk_proj_obs(419, year = 2025)`, while observations for a single month — for example, May — are obtained with `mnk_proj_obs(419, year = 2025, month = 5)`.

```{r project_obs:proj, include = TRUE , echo=TRUE}
prj_obs <- mnk_proj_obs(419, year= 2025, month=5)

prj_obs[2:14]

```


### <u>- User Queries</u>  

User-related functions will be illustrated using the user Xavier Salvador.
  
### ● `mnk_user_byname()`
 
Initially, only an approximate name is known, so a search is performed to retrieve the corresponding user_login. For this example, we start with the query "xavi", which returns a list of possible logins. The desired user — Xavier Salvador — is then selected from that list.

```{r user_name, include = TRUE , echo=TRUE}

user_name <- mnk_user_byname("xavi")

user_name

```


### ● `mnk_user_info()`
 
Once the user ID is known, detailed information can be retrieved with mnk_user_info(). For Xavier Salvador (login "xasalva"), the user ID is 4
   
```{r user_info, include = TRUE , echo=TRUE}
user_info <- mnk_user_info(4)

user_info

```


### ● `mnk_user_proj()`
 
 `mnk_user_proj()` returns the projects to which a user is explicitly subscribed, given the user ID. For Xavier Salvador (user ID 4), the list is obtained with `mnk_user_proj(4)`.

Note: Projects in which a user has contributed observations but is not formally subscribed cannot be retrieved directly in R.
 
```{r user_project, include = TRUE , echo=TRUE}
user_project <- mnk_user_proj(4)

user_project

```


### ● `mnk_user_obs()`
 
Returns all observations for a user within a selected year, and optionally within a specific month. For Xavier Salvador (user ID 4), observations for the entire year are obtained with `mnk_user_obs(4, year = 2025)`, while observations for a single month, for example May, are obtained with `mnk_user_obs(4, year = 2025, month = 5)`.

```{r user_obs, include = TRUE , echo=TRUE}
user_obs <- mnk_user_obs(user_id= 4, year = 2025, month = 8)

user_obs

```


### <u>- Place Queries</u>

Place-related functions will be illustrated using the place  Piscines del Fòrum
 
### ● `mnk_place_byname()`
 
Initially, only an approximate place name is known, so a search is performed to retrieve the corresponding place ID. For this example, we start with the query "forum", which returns a list of possible places. The project actually uses "Piscinas del Forum", which does not match the initial assumption, so it is recommended to try several searches with different terms until the desired place is identified.

```{r place_name, include = TRUE , echo=TRUE}
places <- mnk_place_byname("Forum")

places[,1:6]

```

### ● `mnk_place_sf()`
 
Returns the sf geometry for a place given its place ID. Once the place ID is known, the geometry can be retrieved with `mnk_place_sf()`. For Piscinas del Forum, the geometry is obtained with `mnk_place_sf(253)`.

In the example the geometry is visualized with the leaflet package using the function's default projection, WGS84 (EPSG:4326), the standard used by Google Maps. Note that the returned sf object already includes the place name as an attribute.

```{r place_sf, include = TRUE , echo=TRUE}
# 1. Downloading the geometry 

place <- mnk_place_sf(253)

# 2. Drawing the map

 forum_sf <-leaflet(place) %>%
                addProviderTiles("OpenStreetMap", group = "OSM") %>%
                addProviderTiles("Esri.WorldImagery", group = "Satélite") %>%
                addPolygons(
                  color = "#2c4fb8",
                  weight = 2,
                  opacity = 1,
                  fillOpacity = 0.4,
                  label = ~name, # information added from previous function
                  highlightOptions = highlightOptions(weight = 3, bringToFront = TRUE)
                ) %>%
                addLayersControl(baseGroups = c("Satélite", "OSM")) 
 forum_sf

```

### ● `mnk_places_obs()`

Returns all observations recorded within a place within a selected year, and optionally within a specific month. The result is returned as a data frame without geometry. For Piscines del Fòrum, observations for february  2025 are obtained with `mnk_places_obs(place_id, year = 2025, month =2)`.

Although the output has no geometry, it can be converted to an `sf` object with the helper function `mnk_obs_sf()`. The resulting points can then be mapped with the leaflet package.

```{r places_obs, include = TRUE , echo=TRUE}

obs_place <- mnk_place_obs(place_id = 253, year = 2025, month = 2, quiet = TRUE)

obs_place

#Turning the dataframe into an sf object with the mnk_obs_sf() function

obs_sf <- mnk_obs_sf(obs_place,"id","taxon_name","observed_on", "url_picture", "uri", "user_login")

# Preparing the obtained data for later display in the marker popup

popup_final <-  paste0( "ID: <a href='", obs_sf$uri, 
                        "' target='_blank'>",obs_sf$id , "</a><br>",
                        "Specie:", obs_sf$taxon_name, "<br>",
                        "Observer: ", obs_sf$user_login,"<br>",
                        "Date:", obs_sf$observed_on, "<br>",
                        "<a href= '", obs_sf$url_picture, "' target='_blank'><img src='", 
                        obs_sf$url_picture,
                        "' style='margin-top:2px;border-radius:4px;'> </a> ")
#finally plotins

# Only the observations:

leaflet(obs_sf) %>%
                addProviderTiles("OpenStreetMap", group = "OSM") %>%
                addProviderTiles("Esri.WorldImagery", group = "Satélite") %>%
                addMarkers(lng = ~longitude,
                           lat = ~latitude,
                           popup = popup_final) %>%
                addLayersControl(baseGroups = c("Satélite", "OSM")) 

#  Observations plus place maping

 forum_sf %>%
  addMarkers(data = obs_sf, popup = ~popup_final)

```

### <u>- Observation Queries</u>

### ●`mnk_obs_id()`
 
Returns a single observation given its observation ID. This function is seldom used in practice, as observation IDs are not usually known beforehand.

```{r obs_id, include = TRUE , echo=TRUE}

obs_id <- mnk_obs_id(id = 553028)

obs_id

```

### ● `mnk_obs()`
 
 This is the core function of the package. It is highly versatile and allows searches by project, user, place, date, week number, taxon, or geographic area. Two important notes apply to all `mnk_obs()` calls:

When the query is executed, a message reports the total number of observations found. If this number exceeds 10,000, only the first 10,000 are downloaded by default. To retrieve all records, set `limit_download = FALSE`.

To retrieve only validated records, use `quality = "research"`.Previous examples have already covered searches by project, user and place, so here we focus on two cases not yet shown: a) search by taxon and b) search by bounding box.

a) <u>Taxon search</u>. Observations for the genus Raja and for the species Raja brachyura during 2025 are obtained with `taxon_name = "Raja"` and `taxon_name = "Raja brachyura"` respectively.

```{r obs, include = TRUE , echo=TRUE}
#In this example don´t show messages in console (quiet= TRUE)

# Get observations for genus Raja in 2025 by user Xavier Salvador ( user_id=4)

obs <- mnk_obs(taxon_name = "Raja", year = 2025, user_id = 4, quiet = TRUE, 
               quality = "research")

obs # show full dataframe

# Get observations for species Raja brachyura in 2023 by user Xasalva (user_id= 4)

obs_brachyura <- mnk_obs(taxon_name = "Raja brachyura", year = 2023, user_id = 4, 
                         quiet = TRUE, quality = "research")

obs_brachyura[, c(1,2, 10, 11, 16, 27)]  # show selected columns

```
Images from observations can be visualized and printed using the magick package and the observation's `url_picture` as follow.

```{r brachyura-pic, echo=TRUE, out.width='85%', fig.align='center', fig.alt="Raja undulata observed by Xavier Salvador"}
# Get observations for specie Raja undulata in 2022 by user Xasalva (user_id= 4)

obs_undulata <- mnk_obs(taxon_name = "Raja undulata", year = 2022, user_id = 4, 
                         quiet = TRUE, quality = "research")

# select 3rd observation and get picture URL (url_picture)

url_pic_undulata <- as.character(obs_undulata$url_picture[3])

# read image directly from URL

img_undulata <- image_read(url_pic_undulata )

  # add copyright caption at top
  img_undulata <- image_annotate(
    img_undulata,
    "© Xavier Salvador",
    size = 10, 
    gravity = "northwest", 
    color = "white", 
  ) 

# display image in the document

img_undulata

```

b) <u>Bounding-box search</u>. Observations within a rectangular area are retrieved by supplying the coordinates of the southwest and northeast corners. For example, to search around Platja de Sant Sebastià in Barcelona, define the bounding box and pass it to `mnk_obs()`.

```{r obs_1, include = TRUE , echo=TRUE}
# read shapefile of  study area using relative path 

shp <- "../inst/extdata/espigo_w.shp"

espigo <- sf::st_read(shp, quiet = TRUE)

# ensure WGS84 for leaflet (standard for web maps)

bounds <- st_transform(espigo, 4326)

```
Similarly, the area can also be defined by a vector from two points, Pmax and Pmin, 
where the area is defined as:

`bounds <- c(ymax, xmax, ymin, xmin)`

Applied to the Espigo del W example, this would be:

`bounds <-c(41.371239, 2.193971, 41.368182, 2.189147)`

The two points, pmax and pmin, will be displayed together with the study area
of the Espigo del W in leaflet to help understand how the bounds parameter works.

```{r bounds, include = TRUE , echo=TRUE}

#Pmin
xmin <- 2.189147
ymin <- 41.368182

#Pmax
xmax <- 2.193971
ymax <- 41.371239

pts <- data.frame(
  id = c("Pmin", "Pmax"),
  lng = c(xmin, xmax),
  lat = c(ymin, ymax)
)

# create interactive map for visualising the study area and the two points

bound <- leaflet() %>%
                addProviderTiles("Esri.WorldImagery", group = "Satélite") %>%
                addProviderTiles("OpenStreetMap", group = "OSM") %>%
                addCircleMarkers(             
                         data = pts,
                          radius = 20,
                          color = "cyan", 
                          weight = 2, 
                          fillOpacity = 0.7,
                          label = ~id,          
                          popup = ~paste0("<b>", id, "</b><br>",
                                          "lon: ", round(lng, 6), "<br>",
                                          "lat: ", round(lat, 6))) %>%
                addPolygons(data =    bounds,
                          color = "#2c4fb8",
                          weight = 2,
                          opacity = 1,
                          fillOpacity = 0.4,
                          label = "Bound Espigo W",
                          popup = "Bound Espigo W",
                          highlightOptions = highlightOptions(weight = 3, 
                                                              bringToFront =TRUE)) %>%
                addLayersControl(baseGroups = c("Satélite", "OSM"))

bound  # display map in document

#Obtaining the observations within the study area

obs_torpedo_bounds <- mnk_obs(taxon_name = "Torpedo", year = 2024,
                      bounds = bounds , quality = "research", quiet = TRUE)

obs_torpedo_bounds

#Turning the dataframe into an sf object with the mnk_obs_sf() function and 
#selecting some fields ("id","taxon_name","observed_on",.....) from the original
#data frame for to show later in leaflet popups.

obs_bounds_sf <- mnk_obs_sf(obs_torpedo_bounds,"id","taxon_name","observed_on", "url_picture", "uri", "user_login")

# Preparing the obtained data for later display in the marker popup

popup_final_torpedo <-  paste0( "ID: <a href='", obs_bounds_sf$uri,
                        "' target='_blank'>",obs_bounds_sf$id , "</a><br>",
                        "Specie:", obs_bounds_sf$taxon_name, "<br>",
                        "Observer: ", obs_bounds_sf$user_login,"<br>",
                        "Date:", obs_bounds_sf$observed_on, "<br>",
                        "<a href= '", obs_bounds_sf$url_picture, 
                        "' target='_blank'><img src='", 
                        obs_bounds_sf$url_picture, 
                        "' style='margin-top:2px;border-radius:4px;'> </a> ")

#final plot

bound %>%
  addMarkers(data = obs_bounds_sf, popup = ~popup_final_torpedo)

```


### ● `mnk_obs_byday()`
 
Works like `mnk_obs()` but uses an explicit date interval instead of year or month parameters. The interval is defined by:

`d1`: start date in 'yyyy-mm-dd' format
`d2`: end date in 'yyyy-mm-dd' format

```{r obs_byday, include = TRUE ,echo=TRUE}
#Retrieves observations for specie Raja undulata by user Xasalva (user_id= 4) the observations between 01-01-2024 and  25-05-2024.

obs_undulata_2024 <- mnk_obs_byday(taxon_name = "Raja undulata", d1 = "2024-01-01", 
                             d2= "2024-05-25", quiet = TRUE, quality = "research")

obs_undulata_2024

```


### <u>- Auxiliary functions</u>

The examples below show how the helper functions work. The first two use the *Torpedo torpedo* observations at the Espigó Hotel W from the previous sections. The last two use separate, specific examples.

### ● `mnk_obs_sf()`

This function converts an observations data frame into a spatial layer in sf format.
The function takes the latitude/longitude columns returned by `mnk_obs` or by any other function in the rminka package that returns an observations data frame — such as `mnk_proj_obs` or `mnk_user_obs` — and converts them into an sf POINT layer. By default the geometry is returned in CRS EPSG:4326 (WGS84), but you can specify any CRS you need via the crs argument.
Its basic usage has already been shown briefly in previous sections. Here we develop a detailed example using the *Torpedo torpedo* observations at the espigó Hotel W.

```{r sf_obs, include = TRUE ,echo=TRUE,message=FALSE, warning=FALSE}

obs_utm <- mnk_obs_sf(obs_torpedo_bounds, id, taxon_name,
                      observed_on, user_login, quality_grade, url_picture,
                      uri, crs = 25831)

# The input obs_torpedo_bounds is a data frame with the full set of observation records. The #function keeps only the selected attribute columns (id, taxon_name, observed_on, user_login, #quality_grade, url_picture, uri); these are preserved in the output sf object together with #the generated geometry in obs_utm.

#The crs = 25831 parameter overrides the default EPSG:4326. 
#Here we request the output directly in EPSG:25831 – ETRS89 / UTM zone 31N, 
#which is the official projected CRS for Catalonia. 
#This is useful for distance calculations and for matching local cartography


# Leaflet natively works in Web Mercator (EPSG:3857). To visualize data in UTM, 
#you must change the map's base CRS:

crs25831 <- leafletCRS(
  crsClass = "L.Proj.CRS",
  code = "EPSG:25831",
  proj4def = "+proj=utm +zone=31 +ellps=GRS80 +units=m +no_defs",
  resolutions = c(8192,4096,2048,1024,512,256,128,64,32,16,8,4,2,1,0.5),
  origin = c(0, 9000000)
)

# Because the map is no longer in 3857, you cannot use standard OSM tiles. 
#You need a WMS that serves imagery in the same CRS. 
#Here we use the Institut Cartogràfic i Geològic de Catalunya (ICGC) service.

leaflet(options = leafletOptions(crs25831, minZoom = 0, maxZoom = 14)) %>%
  
      addWMSTiles(
        "https://geoserveis.icgc.cat/servei/catalunya/mapa-base/wms",
        layers = "orto",
        options = WMSTileOptions(format = "image/png", 
                                 transparent = TRUE, version = "1.3.0")) %>%
    
      addCircleMarkers(
        data = obs_utm, 
        radius = 6,
        color = "red",
        fillOpacity = 0.9,
        popup =  paste0( "ID: <a href='", obs_utm$uri,
                            "' target='_blank'>",obs_utm$id , "</a><br>",
                            "Specie:", obs_utm$taxon_name, "<br>",
                            "Observer: ", obs_utm$user_login,"<br>",
                            "Quality:", obs_utm$quality_grade,"<br>",
                            "Date:", obs_utm$observed_on, "<br>",
                            "<a href= '", obs_utm$url_picture, 
                            "' target='_blank'><img src='", 
                            obs_utm$url_picture, 
                            "' style='margin-top:2px;border-radius:4px;'> </a> "))
```

### ● `export_mnk_qgis()`

### ● `get_wrm_tax()`

### ● `shrt_name()`




 
